---
title: "Data Visualization - Peer Assessment 1"
author: "Elsacha"
date: "July 29, 2015"
output:
  html_document:
    theme: cerulean
  pdf_document: default
---
## Motivation
We live in the age of global warming and temperature observations show that temperatures have risen in the past century or so. The provided data contains temperature deviations from the average level which is calculated based on the period from 1951 to 1980. The goal of this visualization is to see if the average temperatures before the reference period were lower (negative deviations) and higher after the reference period (positive deviations).

The heatmap allows to show 3 variables: 

X-axis: Months (Discrete)

Y-axis: Decades (Discrete)

Deviation (Continuous) is shown by different color intensities and numeric values.

```{r echo = FALSE, results="hide"}
# Prepare the data
raw = read.csv("./data/ExcelFormattedGISTEMPDataCSV.csv", na.strings = "****")
#keep data for months only
years = raw[,1:13]
years$Decade = rep(NA, nrow(years))

#add decades
years$Decade = rep(NA, nrow(years))
years[1:10,]$Decade = rep("1880-1889", 10)
years[1:10,]$Year = c(1:10)

years[11:20,]$Decade = rep("1890-1899", 10)
years[11:20,]$Year = c(1:10)

years[21:30,]$Decade = rep("1900-1909", 10)
years[21:30,]$Year = c(1:10)

years[31:40,]$Decade = rep("1910-1919", 10)
years[31:40,]$Year = c(1:10)

years[41:50,]$Decade = rep("1920-1929", 10)
years[41:50,]$Year = c(1:10)

years[51:60,]$Decade = rep("1930-1939", 10)
years[51:60,]$Year = c(1:10)

years[61:70,]$Decade = rep("1940-1949", 10)
years[61:70,]$Year = c(1:10)

years[71:80,]$Decade = rep("1950-1959", 10)
years[71:80,]$Year = c(1:10)

years[81:90,]$Decade = rep("1960-1969", 10)
years[81:90,]$Year = c(1:10)

years[91:100,]$Decade = rep("1970-1979", 10)
years[91:100,]$Year = c(1:10)

years[101:110,]$Decade = rep("1980-1989", 10)
years[101:110,]$Year = c(1:10)

years[111:120,]$Decade = rep("1990-1999", 10)
years[111:120,]$Year = c(1:10)

years[121:130,]$Decade = rep("2000-2009", 10)
years[121:130,]$Year = c(1:10)

years[131:136,]$Decade = rep("2010-2015", 6)
years[131:136,]$Year = c(1:6)

library(reshape2)

years$Decade = as.factor(years$Decade)
years$Year = as.factor(years$Year)

molten = melt(years, id=c("Decade", "Year"))
colnames(molten) = c("Decade", "Year", "Month", "Dev")
molten$Dev = as.numeric(molten$Dev)
molten = molten[!is.na(molten$Dev),]

decs = molten
decs$Year = NULL

decs.avg = tapply(decs$Dev, list(decs$Decade, decs$Month), mean)
df.decs.avg = as.data.frame(decs.avg)

#Make a Period column
df.decs.avg$Decade = unlist(rownames(df.decs.avg))
rownames(df.decs.avg) = NULL

molten2 = melt(df.decs.avg, id=c("Decade"))
colnames(molten2) = c("Decade", "Month", "Avg.Dev")

#sanity check of average deviation to see if calculations are correct
#for January 1880-1889
mean(years[1:10,]$Jan)#=-30.7
head(molten2)[1,]$Avg.Dev #=-30.7 OK
```
 
#### Plot 1: Heatmap - Average Temperature Deviations 1880-2015


```{r echo=FALSE}
library(ggplot2)
molten2$Month = factor(molten2$Month, ordered=TRUE, levels=c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                                                             "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

ggplot(molten2, aes(x = Month, y = Decade))+
  geom_tile(aes(fill = Avg.Dev/100)) +
  scale_fill_gradient(name="Deviation", low="white", high="red") + 
  geom_text(aes(fill = molten2$Avg.Dev/100, label = round(molten2$Avg.Dev/100, 2)), size=2.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Monthly Temperature Deviations in degrees Celsius by Decade")
#ggsave(file="heatmap_with_values_v4.png", width=10, height=10)
```
 
 
The plot confirms the global warming trend. Lighter colors at the bottom of the map represent temperatures in the beginning of observations and correspond to negative deviations from the mean reference level (1951 - 1980). As we move up the heatmap above the reference level, we see how colors become brighter (positive deviations) and become red (increase of approximately 0.75 degrees Celcius compared to the reference and of 1 degree Celcius compared to the beginning of observations in 1880).

#### Plot 2: Barplot - Average Monthly Temperature Deviations by Period compared to reference
```{r echo=FALSE}
months = raw[,1:13]
months$period = rep(NA, nrow(months))

#set period 1 1880-1919
months[1:40,]$period = rep("1880-1919", 40)

#set period 2 1920-1949
months[41:70,]$period = rep("1920-1949", 30)

#set period 3 - Reference period 1950-1979
months[71:100,]$period = rep("1950-1979 Ref", 30)

#set period 4 1980-1999
months[101:120,]$period = rep("1980-1999", 20)

#set period 5 2000 - 2015
months[121:136,]$period = rep("2000-2015", 16)

months$period = as.factor(months$period)


library(reshape2)

molten3 = melt(months, id=c("Year", "period"))
colnames(molten3) = c("Year", "Period", "Month", "Dev")
molten3$Dev = as.numeric(molten3$Dev)
molten3 = molten3[!is.na(molten3$Dev),]

periods = molten3[,2:4]

periods.avg = tapply(periods$Dev, list(periods$Period, periods$Month), mean)
df.periods.avg = as.data.frame(periods.avg)

#Make a Period column
df.periods.avg$Period = unlist(rownames(df.periods.avg))
rownames(df.periods.avg) = NULL

molten4 = melt(df.periods.avg, id=c("Period"))
colnames(molten4) = c("Period", "Month", "Avg.Dev")

suppressWarnings(print(
ggplot(molten4, aes(x=Period, y=Avg.Dev/100)) +
  geom_bar(stat="identity", color="black", fill="#FF4719") +
  facet_wrap(~ Month, ncol = 3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Decade")+
  ylab("Average Deviation in degrees Celsius")+
  ggtitle("Temperature Deviation by Month and by Period, 1880 - 2015")
))
#ggsave(file="barplot_v4.png", width=7, height=7)

```


The barplot also confirms that temperatures before the reference period were lower (negative deviations) and higher after the reference period (positive deviations).




